Cross-validation of regression models: mixed model examples (edited)

### 2023-10-10 scratch/gm_addition_to_cv-edited.Rmd
add edited version of mixed-models vignette section

---
title: "Cross-validation of regression models: mixed model examples"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validation of regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

.opts <- options(digits = 5)
```

```{r packages, include=FALSE}
library(magrittr)
library(bibtex)
library(kableExtra)
library(latex2exp)
library(cv)
library(car)
library(lattice)
library(latticeExtra)
layer <- latticeExtra::layer  # to mask ggplot2::layer
library(lme4)
library(nlme)
knitr::opts_chunk$set(comment='    ')
```
The following artificial data set exemplifies a difference between cluster-based
and individual-based cross-validation in hierarchical models.

@@

```{r data}
{
  # Generate data set

  # Parameters

  set.seed(123)
  Nb <- 10       # number of groups
  Nw <- 5        # number of individuals within groups
  Bb <- 0        # between-group regression coefficient on group mean
  SDre <- 2.0    # between-group SD of random level relative to group mean of X

  SDwithin <- 0.5    # within group SD

  Bw <- 1       # within group effect of X

  dd <- data.frame(
    group = factor(rep(1:Nb, each = Nw)),
    x = 20 + seq(1, Nb*Nw)
  ) %>%
    within(
      {
        xm  <- ave(x, group, FUN = mean) # within-group mean
        y <- 10 +
          Bb * xm +                  # contextual effect
          Bw * (x - xm) +            # within-group effect
          SDre * rnorm(Nb)[group] +  # random level by group
          SDwithin * rnorm(Nb*Nw)    # random error within groups
      }
    )
}
```

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set with 10 clusters"
pl <- xyplot(y ~ x, dd, groups = group,
             ylim = c(0,20),
             par.settings = list(superpose.symbol = list(pch = 16, cex = 1))) +
    latticeExtra::layer(panel.ellipse(...,center.cex =0))
pl
pl <- update(pl, par.settings = list(superpose.symbol = list(pch = 1, cex = .7)))
```

@@

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept only models

fit0 <- lmer(y ~ 1 + (1|group), dd)
fit0l <- lme(y ~ 1, dd, random = ~ 1|group)

# effect of x and random intercept

fit1 <- lmer(y ~ x + (1|group), dd)
fit1l <- lme(y ~ x, dd, random = ~ 1|group)

# effect of x, mean of x and random intercept

fit2 <- lmer(y ~ x + xm + (1|group), dd)
fit2l <- lme(y ~ x + xm, dd, random = ~1|group)

modlist <- list(fit0, fit0l, fit1, fit1l, fit2, fit2l)
names(modlist) <- c('lmer:~1','lme:~1','lmer:~x','lme:~x','lmer:~x+xm','lme:~x+xm')
modlist <- do.call(models, modlist)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
tbl <- do.call(compareCoefs, modlist)
colnames(tbl) <- names(modlist)
tbl[] <- round(tbl,4)
tbl[is.na(tbl)] <- ''
kbl(tbl, align = 'r', caption = "Comparison of estimated fixed effects")
```

@@

```{r}
dd$fit_mod0__lmer.fe <- predict(fit0, re.form = ~ 0)
dd$fit_mod0__lmer.re <- predict(fit0)
dd$fit_mod0__lme.fe <- predict(fit0l, level = 0)
dd$fit_mod0__lme.re <- predict(fit0l, level = 1)

dd$fit_mod1__lmer.fe <- predict(fit1, re.form = ~ 0)
dd$fit_mod1__lmer.re <- predict(fit1)
dd$fit_mod1__lme.fe <- predict(fit1l, level = 0)
dd$fit_mod1__lme.re <- predict(fit1l, level = 1)

dd$fit_mod2__lmer.fe <- predict(fit2, re.form = ~ 0)
dd$fit_mod2__lmer.re <- predict(fit2)
dd$fit_mod2__lme.fe <- predict(fit2l, level = 0)
dd$fit_mod2__lme.re <- predict(fit2l, level = 1)

dl <- stats::reshape(dd, direction = 'long', sep = '.', timevar = 'effect', varying = grep('\\.', names(dd)))
dl$id <- 1:nrow(dl)
dl <- stats::reshape(dl, direction = 'long', sep = '__', timevar = 'method',  varying = grep('__', names(dl)))
dl$id <- 1:nrow(dl)
dl <- stats::reshape(dl, direction = 'long', sep = '_', timevar = 'modelcode',  varying = grep('_', names(dl)))

dl$model <- factor(c('~ 1', '~ 1 + x', '~ 1 + x + xm')[match(dl$modelcode, c('mod0','mod1','mod2'))])

```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values using lmer with random intercept model"

(pl +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod0' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod0' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3)
) %>% update(
  key=list(
    space = 'right',
    text = list(c('fixed effects only','fixed and random')),
    lines = list(lty=c(1,3))))

```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values using lmer with random intercept model and x as a predictor"
(pl +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod1' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod1' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3)
) %>% update(
  ylim = c(-20,40),
  key=list(
    space = 'right',
    text = list(c('fixed effects only','fixed and random')),
    lines = list(lty=c(1,3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values using lmer with random intercept model and x and the group mean of x as predictors"
(pl +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod2' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod2' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3)
) %>% update(
  ylim = c(0,20),
  key=list(
    space = 'right',
    text = list(c('fixed effects only','fixed and random')),
    lines = list(lty=c(1,3))))
```

@@

```{r cross-validation}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation using lmer and lme comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."

cvs_cases <- cv(modlist, data = dd)
cvs_clusters <- cv(modlist, data = dd, cluster = 'group')

cvs_cases <- cvs_cases %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame
cvs_clusters <- cvs_clusters %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame

cvs_cases$modlist <- names(modlist)
cvs_clusters$modlist <- names(modlist)
cvs_cases$cv_method <- 'cases'
cvs_clusters$cv_method <- 'clusters'

dout <-
  within(
    as.data.frame( rbind( cvs_cases, cvs_clusters)),
    {
      model <- factor(sub('lme.*:','',modlist))
      method <- factor(sub(':.*$','',modlist))
      CV <- as.numeric(as.character(`CV crit`))
      cv_method <- factor(cv_method)
    }
)

dout <- dout[order(dout$model),]

xyplot(CV ~ model , dout, groups = cv_method:method,
       par.settings =
         list(superpose.line = list(lty=c(2,3),lwd = 3),superpose.symbol=list(pch=21:24, fill='transparent', cex =1)),
       type = 'b',
       auto.key = T,
       scales = list(y = list(log=T)))

```

>>>>

---
title: "Cross-validation of regression models: mixed model examples (edited)"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validation of regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

.opts <- options(digits = 5)

knitr::opts_chunk$set(comment="    ")
```

```{r packages, include=FALSE}
library(cv)
library(car)
library(lattice)
library(latticeExtra)
library(lme4)
```
The following artificial data set exemplifies a difference between cluster-based
and individual-based cross-validation in hierarchical models.

@@

```{r data}
# Generate data set

# Parameters:
set.seed(123) # set.seed(9693)
Nb <- 10     # number of groups
Nw <- 5      # number of individuals within groups
Bb <- 0      # between-group regression coefficient on group mean
SDre <- 2.0  # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1  # within group effect of x

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = 20 + seq(1, Nb*Nw)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- 10 +
        Bb * xm +                  # contextual effect
        Bw * (x - xm) +            # within-group effect
        SDre * rnorm(Nb)[group] +  # random level by group
        SDwithin * rnorm(Nb*Nw)    # random error within groups
    }
  )
```

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set with 10 clusters"
plot <- xyplot(y ~ x, data=Data, group=group,
               ylim=c(0, 20),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot
```

@@

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept only model:
fit0 <- lmer(y ~ 1 + (1|group), Data)

# effect of x and random intercept:
fit1 <- lmer(y ~ x + (1|group), Data)

# effect of x, mean of x, and random intercept:
fit2 <- lmer(y ~ x + xm + (1|group), Data)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
compareCoefs(fit0,  fit1, fit2)
```

@@

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(fit0, re.form = ~ 0)
  fit_mod0.re <- predict(fit0)
  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
  fit_mod1.re <- predict(fit1)
  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
  fit_mod2.re <- predict(fit2)
})

Data_long <- reshape(Data, direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data)))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))

Data_long$model <- factor(c("~ 1", "~ 1 + x", "~ 1 + x + xm")
                   [match(Data_long$modelcode, c("mod0","mod1","mod2"))])
```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values with random intercept model"
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
                         effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
                         effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  key=list(
    space="right",
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values with random intercept model and x as a predictor"
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
                         effect == "fe"), 
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
                         effect == "re"), 
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  ylim=c(-20, 40),
  key=list(
    space="right",
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values with random intercept model and x and the group mean of x as predictors"
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" &

                           effect == "fe"), 
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & 
                         effect == "re"), 
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  ylim = c(0, 20),
  key=list(
    space="right",
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

@@

```{r cross-validation}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
modlist <- models("~ 1"=fit0, "~ x"=fit1, "~ x + xm"=fit2)
cvs_cases <- cv(modlist, data=Data, seed=9693)
cvs_clusters <- cv(modlist, data=Data, cluster="group", seed=4576)

cvs_cases <- cvs_cases |> lapply(rbind) |> 
  do.call(rbind, args = _) |> as.data.frame()
cvs_clusters <- cvs_clusters |> lapply(rbind) |> 
  do.call(rbind, args = _) |> as.data.frame()

cvs_cases$modlist <- names(modlist)
cvs_clusters$modlist <- names(modlist)
cvs_cases$cv_method <- "cases"
cvs_clusters$cv_method <- "clusters"

dout <- within(
  as.data.frame(rbind(cvs_cases, cvs_clusters)),
  {
    model <- factor(modlist)
    CV <- as.numeric(as.character(`CV crit`))
  }
)

dout <- dout[order(dout$model), ]

xyplot(CV ~ model, data=dout, groups=cv_method,
       par.settings =
         list(superpose.line=list(lty=c(2, 3), lwd=3), 
              superpose.symbol=list(pch=15:16, 
                                    cex=1.5)),
       type="b",
       auto.key=TRUE,
       scales=list(y=list(log=TRUE)))
```
